Group polarization replications may be marred by high false discovery rates

### 2024-12-21 GroupPolarizationStatmod/paper/ms.tex
model/methods good enough to move forward

the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.

Unchecked social and political polarization undermines cooperation and political
efficacy \autocite{Mason2018UncivilAgreementBook,Klein2020}.  \emph{Group
polarization} is the name that has been given to a group-level process that leads
to greater opinion extremity in a group as the group deliberates some topic (e.g.,
``Do you approve, disapprove, or neither approve nor disapprove of the current
administration's policies?''). Extremity only increases, though, assuming that the
mean opinion of the group is initially biased towards one extreme or the other
\autocite{Brown1986,Brown2020}. Group polarization is important to understand
rigorously, therefore, since if groups become increasingly extreme over time,
while group opinions are consolidated, this would exacerbate political
polarization between political opposition groups, which is already undermines the
ability of partisans in democratic governments to work together. Dozens of studies
and meta-analyses over decades have apparently established the empirical reality
of group polarization
\autocite{Moscovici1969,Myers1970,Isenberg1986,Sunstein2009,Sunstein2019}
\autocite[Ch. 5]{Brown2000}. We believe group polarization should occur in certain
contexts since more extreme group members are likely also to be more stubborn
\autocite{Guazzini2015,Lewandowsky2019}, which should pull consensus to extremes
\autocite{Turner2020}. However, many or most of the evidence for group
polarization relies on experimental designs that use ordinal measurements that are
then analyzed as if they were continuous. Unfortunately, this procedure is
unfortunately known to induce false positives if it is not properly accounted for
\autocite{Liddell2018}.

@@

The false detection rate is the scaled probability that a detection, $D$, of
an effect, $E$, is in fact a false detection and there is \emph{not} an effect,
denoted $\neg E$. The probability that a detection $D$ has occurred when there
was no effect, i.e. when $\neg E$, is denoted $\alpha = \Pr(D | \neg E)$.
This is known as the Type I or family-wise error rate. To calculate the 
false detection rate ($\fdr$), one must compare $\alpha$ to the statistical power,
the probability of a detection when there is an effect, denoted $W = \Pr(D | E)$.
Both these values must be scaled by the rate at which effects occur, denoted $b$.
Using the notation outlined above, the false detection rate is
\begin{equation}
  \fdr = \frac{(1 - b)\alpha}{(1 - b)\alpha + bW}.
\end{equation}

@@

The measurement process with $N\to\infty$ participants is simulated by 
integrating the latent opinion distribution between ordinal bin thresholds, 
of which there are $B+1$, where $B$ is the number of opinion bins, with sequential
integer values $b$. In the Moscovici and Zavalloni (1969) example, $b \in
\{-3,-2,\ldots,2,3\}$ and $B = 7$.
Bin $b$ is defined by its lower and upper continuous-valued thresholds, 
$\theta_{b-1}$ and
$\theta_{b}$, respectively. Thresholds are given by the following equations
(and visualized as dotted lines in Figure~\ref{fig:distros}).
\begin{equation}
\begin{cases}
  \theta_0 &= -\infty \\
  \theta_{B} &= \infty \\
  \theta_b &= b + \frac{1}{2} \text{ if } 0 < b < B.
\end{cases}
\end{equation}
\noindent
With thresholds defined, the frequency of ordinal measurment $b$ at time $t$ 
is (Figure~\ref{fig:distros})
\begin{equation}
  p(b;~\mu_t, \sigma_t, B) = \int_{\theta_{b-1}}^{\theta_b} p(o; \mu_t, \sigma_t)
  do.
  \label{eq:ordinal-frequency}
\end{equation}

@@

    \Require $\mu$, $\sigma$, $\bar{\mathbf{o}}$ 
      \Comment{Static guess for common mean, initial guess of variance.}

@@

    \State $\Delta \gets \meanobs(\mu, \sigma) - \meanobsemp$ 
      \Comment{Initialize simulation error.}
    \While {$\Delta^2 > \epsilon$ and $i < i_\mathrm{max}$} \Comment{Search until desired
    accuracy or max iterations reached.}
      \State $\sigma' \sim \mathcal{N}(\sigma, \delta)$ \Comment{Draw new variance to test.}
      \State $\Delta' \gets \bar{o}(\mu, \sigma') - \bar{\mathbf{o}}$
      \If {$\Delta' - \Delta < 0$}
        \State $\sigma \gets \sigma'$
          \Comment{Update $\sigma$ if the error is less than before.}
      \EndIf
      \State $\Delta \gets \bar{o}(\mu, \sigma) - \bar{\mathbf{o}}$
        \Comment{Update simulation error if $\sigma$ is updated.}
    \EndWhile \\

@@

\subsection{Calculate false detection rate from simulations}

False detection rates are calculated using Cohen's $d$ to measure effect
size, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
\end{equation}. 
\noindent
Denote the collection of fitted parameters from the $i^\mathrm{th}$ simulation trial
for experimental condition $c$ and journal article $a$ as
% $\thetafit = \left{\mu_t = \left{\mupre,~\mupost\right},~\sigma_t = \left{\mupre,~\mupost\right}\right}$.
$\thetafit_{a,c,i} = \{\mupre,~\mupost,~\mupre,~\mupost\}_{a,c,i}$.
These parameters are the modal estimates for each parameter obtained from
an ordered probit statistical model fit with
Bayesian Markov-Chain Monte Carlo (MCMC)~\cite[p. 338]{Liddell2018}.

For each $a,c$, we calculate $\alpha_{a,c}(d^*) =
\Pr(D |\neg E;~a,~c,~d^*)$ for a given significance value, $d^*$, directly as
the fraction of $d$ values that are greater than $d^*$.

To calculate article-level false detection curves we calculate the average
across , we set $\alpha_{a,c} = 0.0$ for those
conditions for which we failed to generate spurious group polarization in the
large-$N$ model.

@@

>>>>

Unchecked social and political polarization undermines cooperation and
political efficacy \autocite{Mason2018UncivilAgreementBook,Klein2020}.
\emph{Group polarization} is the name given the social phenomenon of 
when a group becomes more extreme in mean opinion over
time, assuming the mean opinion is biased to one extreme to begin with.
\mt{now hit them with the real shit to intro the problem, our analysis, and
how this can be used constructively going forward.}

\mt{Seems like a good second paragraph structurally. We should give more
details about the scope of the study, group polarization and common
elements of experimental designs.} 
Group polarization experiments tend to employ a common set of design
elements incorporated in our computational model analysis. Group
polarization experiments mostly use ordinal (see exceptions
\cite{Blascovich1974,Blascovich1975,Blascovich1976,Keating2016}). The opinions (e.g., ``Do you approve,
disapprove, or neither approve nor disapprove of the current
administration's policies?''). Extremity only increases, though, assuming
that the mean opinion of the group is initially biased towards one extreme
or the other \autocite{Brown1986,Brown2020}. Group polarization is important
to understand rigorously, therefore, since if groups become increasingly
extreme over time, while group opinions are consolidated, this would
exacerbate political polarization between political opposition groups, which
is already undermines the ability of partisans in democratic governments to
work together. Dozens of studies and meta-analyses over decades have
apparently established the empirical reality of group polarization
\autocite{Moscovici1969,Myers1970,Isenberg1986,Sunstein2009,Sunstein2019}
\autocite[Ch. 5]{Brown2000}. We believe group polarization should occur in
certain contexts since more extreme group members are likely also to be more
stubborn \autocite{Guazzini2015,Lewandowsky2019}, which should pull
consensus to extremes \autocite{Turner2020}. However, many or most of the
evidence for group polarization relies on experimental designs that use
ordinal measurements that are then analyzed as if they were continuous.
Unfortunately, this procedure is unfortunately known to induce false
positives if it is not properly accounted for \autocite{Liddell2018}.

@@

The false detection rate is the scaled probability that a detection, $D$, of
an effect, $E$, is in fact a false detection and there is \emph{not} an effect,
denoted $\neg E$. The probability that a detection $D$ has occurred when there
was no effect, i.e. when $\neg E$, is denoted $\alpha = \Pr(D | \neg E)$.
This is known as the Type I or family-wise error rate. To calculate the 
false detection rate ($\fdr$), one must compare $\alpha$ to the statistical power,
the probability of a detection when there is an effect, denoted $W = \Pr(D | E)$.
Both these values must be scaled by the rate at which effects occur, denoted $b$.
Using the notation outlined above, the false detection rate is
\begin{equation}
  \fdr = \frac{(1 - b)\alpha}{(1 - b)\alpha + bW}.
  \label{eq:fdr}
\end{equation}

@@

The measurement process with $N\to\infty$ participants is simulated by 
integrating the latent opinion distribution between ordinal bin thresholds, 
of which there are $B+1$, where $B$ is the number of opinion bins, with sequential
integer values $b$. In the Moscovici and Zavalloni (1969) example, $b \in
\{-3,-2,\ldots,2,3\}$ and $B = 7$.
Bin $b$ is defined by its lower and upper continuous-valued thresholds, 
$\theta_{b-1}$ and
$\theta_{b}$, respectively. Thresholds are given by the following equations
(and visualized as dotted lines in Figure~\ref{fig:distros}).
\begin{equation}
\theta_b = \begin{cases}
  -\infty         & \text{ if } b=0 \\
  \infty          & \text{ if } b=B \\
  b + \frac{1}{2} & \text{ otherwise.}
\end{cases}
\end{equation}
\noindent
With thresholds defined, the frequency of ordinal measurment $b$ at time $t$ 
is (Figure~\ref{fig:distros})
\begin{equation}
  p(b;~\mu_t, \sigma_t, B) = \int_{\theta_{b-1}}^{\theta_b} p(o; \mu_t, \sigma_t)
  do.
  \label{eq:ordinal-frequency}
\end{equation}

@@

    \Require $\mu$, $\sigma$, $\bar{\mathbf{o}}$ 
      \Comment{Set latent mean, guess variance that yields, input empirical mean.}

@@

    \State $\Delta \gets \meanobs(\mu, \sigma) - \meanobsemp$ 
      \Comment{Initialize simulation error.}
    \While {$\Delta^2 > \epsilon$ and $i < i_\mathrm{max}$} \Comment{Search until desired
    accuracy or max iterations reached.}
      \State $\sigma' \sim \mathcal{N}(\sigma, \delta)$ \Comment{Draw new variance to test.}
      \State $\Delta' \gets \bar{o}(\mu, \sigma') - \bar{\mathbf{o}}$ 
        \Comment{Calculate error using new variance.}
      \If {$\Delta' - \Delta < 0$}
        \State $\sigma \gets \sigma'$
          \Comment{Update $\sigma$ if the error is less than before.}
      \EndIf
      \State $\Delta \gets \bar{o}(\mu, \sigma) - \bar{\mathbf{o}}$
        \Comment{Update simulation error if $\sigma$ is updated.}
    \EndWhile \\

@@

\subsection{Outcomes: Estimating and limiting false detection rates}

To calculate the false detection rate we begin by simulating 
pre- and post-deliberation 
opinions by drawing $N_e$ samples from normal latent opinion
distributions that generated spurious group polarization, i.e., with
parameters labelled $\beta_e$ found in the hillclimbing step. Binning these
random values simulates observations. $N_e$ is the number
of participants in experimental condition $e$. Each trial $i$ of $N_e$
simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model. 
The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
\end{equation}
\noindent
The family-wise error rate is, $\alpha = \Pr(D|\neg E)$, is the 
fraction of trials with $d > d^*$. To calculate the family-wise error rate
for a study, $s$ which contains several experimental conditions, $e$,
\begin{equation}
  \alpha_s = \frac{1}{|s|} \sum_{e \in s} \alpha_e
\end{equation}
\noindent
The grand mean across all studies for the corpus is 
\begin{equation}
  \alpha_\mathrm{all} = \frac{1}{|E|} \sum_{e \in E} \alpha_e.
\end{equation}
With $\alpha$ calculated, we calculate the false discovery rate using
Equation~\ref{eq:fdr}.

We invert this procedure to find what significance value $d^*$ is necessary
to achieve a low family-wise error rate $\alpha_e = 0.05$ for experimental
condition, $e$. To do this we first calculate $\alpha_e(d^*)$ for a
sufficiently wide range of $d^*$. We then find the lowest $d^*$ for which
$\alpha_e(d^*) \leq 0.05$.

@@

Our analysis brings some additional questions into focus. Most
significantly, recall that even the commensurate ordered probit Bayesian
model often failed to accurately detect negative results. A more rigorous
analysis would include additional sources of variance that will likely
further inflate the family-wise error rate, $\alpha$, and the false
discovery rate~\cite{Yarkoni2022}. First and foremost, there is variance
within each experimental deliberation group; no group polarization study
accounts for this, instead treating participants in each experimental
condition as if they were one large group. Furthermore, survey data are
known to be noisy~\cite{Zaller1992}. People report opinions differently
over time for no apparent reason (i.e., opinions are \emph{unstable}).
Context matters: the order in which a survey question is asked, and which
question framing is chosen among logically equivalent alternatives, can
both significantly influence participant responses. The effect of 
accounting for these sources of variance must be understood to estimate
$\alpha$ (and power, $W$) for the design of the next generation of group
polarization experiments.
